\textbf{Department Invited Speakers Do Not Reflect Trainee Diversity}

### 2019-09-11 submission/manuscript.Rmd
work on discussion and tool description, draft for comments

---
title: '**Department Invited Speakers Do Not Reflect Trainee Diversity**'
csl: mbio.csl
fontsize: 10pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
geometry: margin=1.0in

---

@@

Several papers have investigated the representation of women at scientific conferences, however, we have only identified one that focused on invited speakers at universities (Nittrouer, 2018). In their study, Nittrouer et. al., examined 3,652 talks at 50 U.S. institutions in 2013 - 2014 and found that women faculty are less likely to be invited speakers, despite similar acceptance rates.

These results suggest that women faculty are less likely to invited as speakers, a decision that may be negatively impacted by assumptions about competency and dedication. The dedication of women who have children to their work is perceived to be less than that of their colleagues, e.g., men who also have children. The perceived prioritization and commitments of women to family over work may cause faculty to doubt their acceptance of a speaking invitation (despite the prestigious nature of these invitations and evidence to the contrary), causing the faculty member to invite a different colleague who they feel is more likely to agree. Departments have different processes and criteria for selecting invited speakers, but want to bring the best scientists possible. It may also be that the definition of "best" poses a problem to women, who need three-times as many publications as their men colleges to be considered equally competent. Some departments only invite tenured faculty, which severely limits the number of potential women speakers. Alternately, pre-tenure faculty members invite prestigious, tenured faculty in their field to network and secure letters for their own tenure package. The increased burden of women to prove competency decreases their likelihood to be considered for either tenure and as possible source of tenure letters.

+ studies that go beyond gender -- e.g., race/ethnicity (this seems to be the first)

## Call to Action
+ Improving speaker diversity
    + US-serving institutions have particular responsibility to those historically suppressed populations
    + to improve retention of white women & HURMs, each group needs equivalent representation to counteract biases and improve self-efficacy
    + recall bias not the only issue, also not always possible to identify members of historically under-served communities
    + example: there are speakers who have URM status, but it wasn't readily apparent from the internet/CV 
    + perspective is often as/more important than self-identification

+ Development of Diversify tools
    + inspired by EEB/Chemistry
    + tool for members of URM groups to self-identify & for others to use to find diverse candidates
    + describe maintenance of lists (Rebecca)
    + describe website creation

+ Other resources/ideas
    + student/lab invited speakers -- improve diversity of suggestions (fields/careers)
    + departments to invite speakers to share about their personal story as well as their science

@@

+ increased retention of white women & HURMS -- increasing representation of 
+ tools available to create field specific lists of H under-served

>>>>

---
title: '**Department Invited Speakers Do Not Reflect Trainee Diversity**'
csl: mbio.csl
fontsize: 10pt
output:
  word_document: default
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
geometry: margin=1.0in
---

@@

Several papers have investigated the representation of women at scientific conferences, however, we have only identified one that focused on invited speakers at universities (Nittrouer, 2018). In their study, Nittrouer et. al., examined 3,652 talks at 50 U.S. institutions in 2013 - 2014 and found that women faculty are less likely to be invited speakers, despite similar acceptance rates. These results suggest that women faculty are less often invited as speakers, a decision that may be negatively impacted by assumptions about competency and dedication. The dedication of women who have children to their work is perceived to be less than that of their colleagues, i.e., men who also have children. The perceived prioritization and commitments of women to family over work may cause faculty to doubt their acceptance of a speaking invitation (despite the prestigious nature of these invitations and evidence to the contrary). As a result, the faculty member invites a different colleague who they feel is more likely to agree (and is a man). Departments have different processes and criteria for selecting invited speakers, but it is a matter of pride to bring the best scientists possible. It may be that the definition of "best" poses a problem to women, who need three-times as many publications as their men colleges to be considered equally competent. Some departments only invite tenured faculty, which severely limits the number of potential women speakers. Another scenario is that pre-tenure faculty members invite prestigious, tenured faculty in their field to network and secure letters for their own tenure package. The increased burden of women to prove competency decreases their likelihood to be considered for either tenure or as possible source of tenure letters.

We have not been able to identify any other publications examining scientific speaker diversity beyond gender. This seems to be the first. This is concerning since conclusions drawn from gender-based studies are often framed, and considered, to be applicable to other marginalized groups (e.g., HURM). This is a flawed assumption. While there is no doubt some overlap, each group remains marginalized due to a complex set of factors that are unique to each group and cannot always be solved by gender-based solutions. US-serving institutions, such as the University of Michigan have a particular responsibility to the historically suppressed populations included in our definition of HURMs. We therefore implore US institutions to apply this framing to their discussions and research.

Our data support the advancement of social role theory, which states that to improve retention of white women & HURMs, each group needs equivalent representation to counteract biases and improve self-efficacy. Implict biases that affect perceptions of marginalized groups are the primary issue, but we must acknowledge that it is not always possible to identify members of historically under-served communities. For instance, after data analysis, we learned that at least one speaker in our data set is a member of a HURM group, but it wasn't readily apparent from their internet presence or CV. This limitation makes two important points: that perspective is often as, if not more, important than self-identification with regards to biased outcomes, and that we need better tools to identify members of marginalized groups.

## Building Diversify

To help address this issue, we make some suggestions (Table 1) and have developed a resource to identify scientists who are members of marginalized and/or historically under-served groups. Motivated by a lack of such resources and inspired by similar resources--DiversifyEEB and DiversifyChemistry--we created DiversifyMicrobiology and DiversifyImmunology. These resources are a tool for symposium organziers, award committees, search committees and other scientists to identify individuals to diversify their pools. Additionally, we have built these as a tool for use by other fields and organizations to create their own lists. Importantly, since these lists are compiled by self-nomination, we can ensure that only scientists comfortable revealing their marginalized identities are included.

+ Development of Diversify tools
    + describe maintenance of lists (Rebecca)

The website provides an interface to the Google forms and spreadsheets with template pages for viewing the list, adding a name to the list, and finding additional resources. Importantly, our website creation tool is hosted for free by GitHub, which provides a free website for each GitHub organization (citation). Basic tools and skills required to set up a Diversify site include knowledge of, or experience with, the version control tool git, the webtool GitHub, and a text editor. A tutorial in the DiversifyMicrobiology repository on GitHub provides links to these resources and instructions for adapting the tool to your own field.

@@

To increase the retention of white women and HURMS in STEM, they must also be represented as experts. However, the invited speaker diversity at one department does not represent the diversity of trainees. There is a lack of research on invited speakers examining factors other than gender. To facilite the identification and recruitment of individuals in these historically under-served groups, we have built a tool to create self-nominated, field-specific lists.

### 2019-09-11 submission/manuscript.tex
work on discussion and tool description, draft for comments

Several papers have investigated the representation of women at
scientific conferences, however, we have only identified one that
focused on invited speakers at universities (Nittrouer, 2018). In their
study, Nittrouer et. al., examined 3,652 talks at 50 U.S. institutions
in 2013 - 2014 and found that women faculty are less likely to be
invited speakers, despite similar acceptance rates.

These results suggest that women faculty are less likely to invited as
speakers, a decision that may be negatively impacted by assumptions
about competency and dedication. The dedication of women who have
children to their work is perceived to be less than that of their
colleagues, e.g., men who also have children. The perceived
prioritization and commitments of women to family over work may cause
faculty to doubt their acceptance of a speaking invitation (despite the
prestigious nature of these invitations and evidence to the contrary),
causing the faculty member to invite a different colleague who they feel
is more likely to agree. Departments have different processes and
criteria for selecting invited speakers, but want to bring the best
scientists possible. It may also be that the definition of ``best''
poses a problem to women, who need three-times as many publications as
their men colleges to be considered equally competent. Some departments
only invite tenured faculty, which severely limits the number of
potential women speakers. Alternately, pre-tenure faculty members invite
prestigious, tenured faculty in their field to network and secure
letters for their own tenure package. The increased burden of women to
prove competency decreases their likelihood to be considered for either
tenure and as possible source of tenure letters.

\subsection{Call to Action}\label{call-to-action}

\begin{itemize}
\tightlist
\item
  Improving speaker diversity

  \begin{itemize}
  \tightlist
  \item
    US-serving institutions have particular responsibility to those
    historically suppressed populations
  \item
    to improve retention of white women \& HURMs, each group needs
    equivalent representation to counteract biases and improve
    self-efficacy
  \item
    recall bias not the only issue, also not always possible to identify
    members of historically under-served communities
  \item
    example: there are speakers who have URM status, but it wasn't
    readily apparent from the internet/CV
  \item
    perspective is often as/more important than self-identification
  \end{itemize}
\item
  Development of Diversify tools

  \begin{itemize}
  \tightlist
  \item
    inspired by EEB/Chemistry
  \item
    tool for members of URM groups to self-identify \& for others to use
    to find diverse candidates
  \item
    describe maintenance of lists (Rebecca)
  \item
    describe website creation
  \end{itemize}
\item
  Other resources/ideas

  \begin{itemize}
  \tightlist
  \item
    student/lab invited speakers -- improve diversity of suggestions
    (fields/careers)
  \item
    departments to invite speakers to share about their personal story
    as well as their science
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  increased retention of white women \& HURMS -- increasing
  representation of
\item
  tools available to create field specific lists of H under-served
\end{itemize}

>>>>

Several papers have investigated the representation of women at
scientific conferences, however, we have only identified one that
focused on invited speakers at universities (Nittrouer, 2018). In their
study, Nittrouer et. al., examined 3,652 talks at 50 U.S. institutions
in 2013 - 2014 and found that women faculty are less likely to be
invited speakers, despite similar acceptance rates. These results
suggest that women faculty are less often invited as speakers, a
decision that may be negatively impacted by assumptions about competency
and dedication. The dedication of women who have children to their work
is perceived to be less than that of their colleagues, i.e., men who
also have children. The perceived prioritization and commitments of
women to family over work may cause faculty to doubt their acceptance of
a speaking invitation (despite the prestigious nature of these
invitations and evidence to the contrary). As a result, the faculty
member invites a different colleague who they feel is more likely to
agree (and is a man). Departments have different processes and criteria
for selecting invited speakers, but it is a matter of pride to bring the
best scientists possible. It may be that the definition of ``best''
poses a problem to women, who need three-times as many publications as
their men colleges to be considered equally competent. Some departments
only invite tenured faculty, which severely limits the number of
potential women speakers. Another scenario is that pre-tenure faculty
members invite prestigious, tenured faculty in their field to network
and secure letters for their own tenure package. The increased burden of
women to prove competency decreases their likelihood to be considered
for either tenure or as possible source of tenure letters.

We have not been able to identify any other publications examining
scientific speaker diversity beyond gender. This seems to be the first.
This is concerning since conclusions drawn from gender-based studies are
often framed, and considered, to be applicable to other marginalized
groups (e.g., HURM). This is a flawed assumption. While there is no
doubt some overlap, each group remains marginalized due to a complex set
of factors that are unique to each group and cannot always be solved by
gender-based solutions. US-serving institutions, such as the University
of Michigan have a particular responsibility to the historically
suppressed populations included in our definition of HURMs. We therefore
implore US institutions to apply this framing to their discussions and
research.

Our data support the advancement of social role theory, which states
that to improve retention of white women \& HURMs, each group needs
equivalent representation to counteract biases and improve
self-efficacy. Implict biases that affect perceptions of marginalized
groups are the primary issue, but we must acknowledge that it is not
always possible to identify members of historically under-served
communities. For instance, after data analysis, we learned that at least
one speaker in our data set is a member of a HURM group, but it wasn't
readily apparent from their internet presence or CV. This limitation
makes two important points: that perspective is often as, if not more,
important than self-identification with regards to biased outcomes, and
that we need better tools to identify members of marginalized groups.

\subsection{Building Diversify}\label{building-diversify}

To help address this issue, we make some suggestions (Table 1) and have
developed a resource to identify scientists who are members of
marginalized and/or historically under-served groups. Motivated by a
lack of such resources and inspired by similar resources--DiversifyEEB
and DiversifyChemistry--we created DiversifyMicrobiology and
DiversifyImmunology. These resources are a tool for symposium
organziers, award committees, search committees and other scientists to
identify individuals to diversify their pools. Additionally, we have
built these as a tool for use by other fields and organizations to
create their own lists. Importantly, since these lists are compiled by
self-nomination, we can ensure that only scientists comfortable
revealing their marginalized identities are included.

\begin{itemize}
\tightlist
\item
  Development of Diversify tools

  \begin{itemize}
  \tightlist
  \item
    describe maintenance of lists (Rebecca)
  \end{itemize}
\end{itemize}

The website provides an interface to the Google forms and spreadsheets
with template pages for viewing the list, adding a name to the list, and
finding additional resources. Importantly, our website creation tool is
hosted for free by GitHub, which provides a free website for each GitHub
organization (citation). Basic tools and skills required to set up a
Diversify site include knowledge of, or experience with, the version
control tool git, the webtool GitHub, and a text editor. A tutorial in
the DiversifyMicrobiology repository on GitHub provides links to these
resources and instructions for adapting the tool to your own field.

@@

To increase the retention of white women and HURMS in STEM, they must
also be represented as experts. However, the invited speaker diversity
at one department does not represent the diversity of trainees. There is
a lack of research on invited speakers examining factors other than
gender. To facilite the identification and recruitment of individuals in
these historically under-served groups, we have built a tool to create
self-nominated, field-specific lists.
